---
title: "Painting with R: Crafting the Dutch Voter Portrait"
output: 
  learnr::tutorial:
    progressive: true
runtime: shiny_prerendered
editor_options: 
  chunk_output_type: console
learnr_to_md_options:
  keep_output: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(message = FALSE, warning = FALSE, fig.path = "img/")
library(learnr)
gradethis::gradethis_setup()
tutorial_options(
  exercise.reveal_solution = FALSE
)
```

<style>
.Info {
    background-color: rgb(204, 229, 255);
    border: 1px solid rgb(184, 218, 255);
    color: rgb(0, 64, 133);
    padding: 1em;
    margin: 1em;
}
.Info::before {
  font-weight: bold;
  content: "🎨 Artistic Insight";
}

</style>

## Crafting the Canvas with CBS and Voting Data

Imagine you're an artist, and you're about to paint a detailed portrait of the Dutch voter.
To do this, you need the right colors and brushes.
In our R world, these are data and functions. 

For our masterpiece, we'll be using vibrant colors from CBS (the Dutch statistics bureau) and the election board's palette.
While you can mix these colors yourself by downloading them, we've prepared a special set for you to make your artistic journey smoother.

Remember, just like an artist uses different brushes and techniques to bring a painting to life, in `R`, we use different functions and commands to visualize and analyze data.
One of the most common formats we'll be using is the `csv` file, which is like a sketchbook where data is neatly organized, ready to be painted upon.

### Laying Down the Base with Demographics Data

When you run the code below, think of it as preparing your canvas.
You'll see a *data frame* - this is like a sketch on your canvas.
In this sketch, each row is a unique Dutch municipality (*gemeente*), and the columns are different shades of demographic information, just like colors on a palette.
Let's have a look at the colors (the [CBS regional figures](https://opendata.cbs.nl/statline/#/CBS/nl/dataset/70072ned/table?ts=1691399765879)):

```{r datademo, exercise=T}
library(tidyverse)
demographics <- read_csv("https://raw.githubusercontent.com/vanatteveldt/ccslearnr/master/data/dutch_demographics.csv")
demographics
```

```{r, include=FALSE}
# Somehow shinyapps.io does not always pick up all the dependencies
library(ggcorrplot)
library(xkcd)
library(sjPlot)
```

Take a moment to admire your canvas.
Can you guess what each shade or column represents?
Click 'continue' to unveil the colors of our palette.

::: Info ::::::
Remember, this is your canvas, and there's no fear of making a mistake.
Every time you feel like starting over, just hit the 'Run code' button, and you'll have a fresh canvas.
If you ever feel you've mixed the wrong colors, the `↻ Start Over` button will bring back the original shades.
:::

### Unveiling the Colors of Our Palette

| Color (Variable) | Shade (Explanation)                                                         |
| ---------------- | --------------------------------------------------------------------------- |
| gm               | Unique code for each municipality                                           |
| gemeente         | Unique name of each municipality                                            |
| v01_pop          | Total population, like the base color of our painting                       |
| v57_density      | The intensity or population density                                         |
| v43_nl           | Percentage of population without a migration history, our native shade      |
| v122_disposable  | Average disposable income per household, the richness of our color palette  |
| v132_income      | Average standardized income per household, the tone of our color            |
| v142_wealth      | Median household wealth, the depth of our shade                             |
| v153_uitkering   | Percentage of population receiving social benefits, the texture of our paint|
| c_65plus         | Percentage of population older than 65, the vintage shade                   |

### Painting the Voting Landscape

Now, let's add more details to our painting with voting data.
This will be like adding intricate patterns and designs to our canvas. 

`R` is like a versatile brush that can help us refine our artwork.
For instance, the code below paints the voting landscape of a single municipality (Groningen). But what if you wanted to change the perspective?
Try removing the `desc(..)` function from arrange and see how the landscape changes  (change `arrange(desc(votes))` into `arrange(votes)`).

**Artistic Challenge**: Can you modify the code to depict the vibrant voting scene of Amsterdam, ordered by the party names (starting from A)?

```{r datavotes, exercise=TRUE}
library(tidyverse)
results <- read_csv("https://raw.githubusercontent.com/vanatteveldt/ccslearnr/master/data/dutch_elections_2023ps.csv")
results |>
  filter(gemeente == "Groningen") |>
  arrange(desc(votes))
```

```{r datavotes-solution}
library(tidyverse)
results <- read_csv("https://raw.githubusercontent.com/vanatteveldt/ccslearnr/master/data/dutch_elections_2023ps.csv")
results |>
  filter(gemeente == "Amsterdam") |>
  arrange(party)
```

```{r datavotes-code-check}
grade_code()
```

```{r datavotes-hint-1}
# To focus on Amsterdam, use gemeente == "Amsterdam" within the filter() brush stroke.
```

```{r datavotes-hint-2}
# To focus on Amsterdam, use gemeente == "Amsterdam" within the filter() brush stroke.
# Change your perspective by replacing 'votes' with 'party' in the arrange brush stroke, and remove the 'desc(.)' technique.
```

```{r datavotes-hint-3}
# To focus on Amsterdam, use gemeente == "Amsterdam" within the filter() brush stroke.
# Change your perspective by replacing 'votes' with 'party' in the arrange brush stroke, and remove the 'desc(.)' technique.
results |> 
  filter(gemeente == "Amsterdam") |> 
  arrange(party)
```

::: Info ::::::
The code box above is like an art challenge with a specific theme. 
You need to adjust your brush strokes and click the `☑️Submit Artwork` button to showcase your masterpiece.
Don't fret over a few misplaced strokes; art is about expression and learning.
You can always refine your techniques and try again.
If you ever feel lost, click on the `💡 Artistic Tips` for some inspiration.
Remember, while `R` can guide you, it's your creativity that brings the canvas to life.
:::

## Painting with R: Crafting Visual Masterpieces

Just as an artist uses brushes and colors to create beautiful paintings, in the world of `R`, we have a powerful toolkit called `ggplot` to craft visual masterpieces.
While mastering the art of `ggplot` might seem like learning to paint a detailed fresco, with the right guidance and practice, you'll soon be creating visual wonders!

For our next artwork, we'll continue using the colors and brushes from our previous canvas:

```{r data, include=FALSE}
library(tidyverse)
results <- read_csv("https://raw.githubusercontent.com/vanatteveldt/ccslearnr/master/data/dutch_elections_2023ps.csv")
demographics <- read_csv("https://raw.githubusercontent.com/vanatteveldt/ccslearnr/master/data/dutch_demographics.csv")
```

```{r dataprep}
library(tidyverse)
results <- read_csv("https://raw.githubusercontent.com/vanatteveldt/ccslearnr/master/data/dutch_elections_2023ps.csv")
demographics <- read_csv("https://raw.githubusercontent.com/vanatteveldt/ccslearnr/master/data/dutch_demographics.csv")
```

### Crafting Your First Scatter Artwork

Imagine a scatter plot as a canvas filled with stars, each shining brightly, revealing a story.
Let's craft our first scatter artwork to explore the relationship between population density and age.
Do the stars of older age shine brighter in the bustling cities or in the tranquil countryside?

```{r scatter, exercise=T, exercise.setup='data'}
ggplot(
  data = demographics,
  mapping = aes(x = v57_density, y = c_65plus)
) +
  geom_point()
```

Gaze upon your creation!
It seems the stars of older age shine more luminously in less dense municipalities.
Dive into the code above and notice how `ggplot` is like your canvas, where you define your colors (`data = demographics`), brush strokes (`mapping = aes(..)`), and how to place these strokes (`v57_density` dictates the placement on the x, `c_65plus` on the y axis).
The `geom_point` is like a special kind of brush that that makes dots (there are others to make lines or symbols and many more).

Feel adventurous?
Try changing the colors or the size of the stars by replacing variables in the mapping.
Let your creativity flow!

### Crafting a Masterpiece

While our initial artwork was a delightful sketch, let's now craft a detailed masterpiece.
The following code might seem like a complex dance of brush strokes, but with a keen eye, you'll see the rhythm and flow.
It also shows quite well how ggplot works: 
You first construct a basic plot consisting of data, aesthetics, and geoms. 
This generally looks quite nice as R is experienced in painting data (it knows what scales, sizes and colors work okay).
However, you can add many optional elements to the plot to change it into your personal style: from changing the color bars to repositioning the legend.

::: Info ::::::
Remember, lines starting with a `#` are like whispered secrets from the artist, guiding you through the creation process. They're comments that `R` will gently ignore, allowing you to understand the essence of each brush stroke.
:::

```{r scatterpretty, exercise=T, exercise.setup='data'}
# Setting up the canvas with ggplot
ggplot(
  data = demographics,
  mapping = aes(x = v57_density, y = v132_income, size = v01_pop, color = 100 - v43_nl)
) +
  geom_point(alpha = .7) +
  # Adding titles and labels to our artwork
  ggtitle("Contrasts between Urban and Rural Landscapes") +
  xlab("Population Density") +
  ylab("Wealth of the Land") +
  # Choosing our color palette
  scale_color_gradient(low = "skyblue", high = "midnightblue") +
  # Refining the legend to guide our viewers
  scale_size(
    breaks = c(5000, 50000, 500000),
    labels = c("5.000", "50.000", "500.000")
  ) +
  guides(
    size = guide_legend(
      title = "Population", title.position = "top",
      override.aes = list(color = "midnightblue")
    ),
    color = guide_colorbar(
      title = "% with Rich Cultural History",
      title.position = "top",
      direction = "horizontal"
    )
  ) +
  # Choosing a classic theme for our masterpiece
  theme_classic() +
  theme(legend.key = element_blank())
```

Each line in the code above is like a brush stroke, adding depth and detail to our artwork.
Feel free to play around, remove some lines, or change the colors.
What happens when you experiment with different themes or guide information?

::: Info ::::::
In the world of `ggplot`, the appearance of our artwork is shaped by *themes*. These themes define the mood, style, and essence of our visual story.
While we've used the `theme_classic`, there are many other themes waiting to be explored.
For a modern touch, try `theme_linedraw` or for a minimalist approach, `theme_void`. If you're feeling adventurous, dive into the `ggthemes` package and experiment with themes like `ggthemes::theme_economist` or `ggthemes::theme_tufte`.
:::



### Merging Canvases: Combining Data Sources

In the realm of art, sometimes we need to combine different sketches to create a detailed masterpiece.
Similarly, in `R`, we often merge data from various sources to craft intricate visualizations.
Let's embark on this artistic journey and see how we can blend our sketches to depict the Dutch voter landscape.

Imagine you have two sketches - one showcasing election results (`results`) and another portraying demographic details.
To create a detailed artwork, we need to merge these sketches based on a common element, the `gm` (like a unique signature on each sketch).

Let's craft a visualization that depicts the support for the farmer's party BBB against the backdrop of population density:

```{r plot, exercise=T, exercise.setup='data'}
bbb <- filter(results, party == "BBB")
data <- inner_join(bbb, demographics)
ggplot(data = data, mapping = aes(x = log10(v57_density), y = votes, color = party, size = v01_pop)) +
  geom_point(alpha = .5) +
  xlab("Population Density (Artistic Log Scale)") +
  ylab("Support for the Party") +
  scale_color_manual(values = c("BBB" = "green", "Anti-immigration" = "blue")) +
  scale_size(guide = "none") +
  ggtitle(
    "Support Landscape for BBB and Anti-immigration Parties",
    "(Dutch 2023 Artistic Elections; note: size of point relative to logged municipality population)"
  ) +
  xkcd::theme_xkcd() +
  theme(legend.position = "top", legend.title = element_blank())
```

## Crafting Maps: Painting Regional Landscapes

To truly appreciate regional differences, let's paint a map.
This is like using a stencil to guide our brush strokes, ensuring we capture every detail.

### Crafting the Stencil: Shape Files

To paint our map, `R` needs a stencil or a *shape* of the regions (like the municipalities we've been discussing).
Let's fetch a detailed stencil of Dutch regions:

```{r loadshapes, include=FALSE, exercise.setup='data'}
library(sf)
if (!file.exists("sf_nl.rds")) {
  library(tidyverse)
  shapes <- read_rds("https://github.com/vanatteveldt/ccslearnr/raw/master/data/sf_nl.rds")
  write_rds(shapes, "sf_nl.rds")
} else {
  shapes <- read_rds("sf_nl.rds")
}
```


```{r datagis, exercise=TRUE}
library(tidyverse)
library(sf)
shapes <- read_rds("https://github.com/vanatteveldt/ccslearnr/raw/master/data/sf_nl.rds")
shapes
```

With our stencil ready, let's start painting!
Using the `geom_sf` brush, we can paint the regions using the `geom` column from our stencil.
Let's add some color using the province column:

```{r plotshape, exercise=TRUE, exercise.setup='loadshapes'}
ggplot(data = shapes) +
  geom_sf(aes(geometry = geom, fill = provincie))
```

### Merging Stencil and Colors: Combining Shapes and Demographics

Our stencil is ready, but to make our map truly come alive, we need to add colors from our demographic palette.
By merging our stencil with demographic details, we can paint regions with sociologically relevant shades.

For instance, let's paint the population density across municipalities (for this we first `join` data once more):

```{r gisdensity, exercise=TRUE, exercise.setup='loadshapes'}
inner_join(shapes, demographics) |>
  ggplot() +
  geom_sf(aes(geometry = geom, fill = v57_density)) +
  scale_fill_gradient(low = "white", high = "red", guide = "none") +
  ggtitle("Population Density Artwork per Municipality") +
  theme_void()
```

Now, let's compare this with the support landscape for the farmer's party BBB:

```{r gisbbb, exercise=TRUE, exercise.setup='loadshapes'}
bbb <- results |> 
  filter(party == "BBB")
inner_join(shapes, bbb) |> 
  ggplot() +
  geom_sf(aes(geometry = geom, fill = votes)) +
  ggtitle("Support Landscape for BBB") +
  scale_fill_gradient(low = "white", high = "green", name = "Support Shades") +
  theme_void() +
  theme(legend.position = "bottom")
```

Gaze upon your creation!
What tales does this artwork tell about the BBB party's audience? 
Feel inspired?
Try painting similar maps for other parties or shades.
Maybe experiment with a different color palette to set a different mood for your artwork.

## Painting with R: Crafting Statistical Masterpieces

In the world of art, sometimes we need more than just colors and brushes.
We need tools to measure, analyze, and understand the patterns in our artwork.
Similarly, in `R`, we have a powerful set of statistical tools to help us decipher the patterns in our data.
Let's dive into this analytical journey and craft some statistical masterpieces.

### Measuring Patterns: Correlation

Imagine correlation as the rhythm or harmony between two colors in a painting.
When two colors blend seamlessly, they are positively correlated.
When they contrast sharply, they might be negatively correlated.
Let's explore how the demographics harmonize with each other:

```{r cor, exercise=TRUE, exercise.setup='data'}
vars <- select(demographics, v01_pop:c_65plus)
cor(vars, use = "pairwise")
```

To visualize this harmony, we can use the `ggcorrplot` package, which paints a beautiful correlation matrix:

```{r corplot, exercise=TRUE, exercise.setup='data'}
library(ggcorrplot)
vars <- select(demographics, v01_pop:c_65plus)
correlations <- cor(vars, use = "pairwise")
ggcorrplot(correlations, method = "circle")
```


### Crafting Predictive Art: Regression

Imagine regression as predicting the next brush stroke based on the patterns of previous strokes.
Let's see if we can predict the support for the BBB party based on demographic details.
We'll merge our sketches, use the `lm` function to craft our predictive model, and then display our findings in a detailed table (using `tab_model` from `sjPlot`):

```{r lm, exercise=TRUE, exercise.setup='data'}
library(sjPlot)
bbb <- filter(results, party == "BBB")
data <- inner_join(bbb, demographics)
m <- lm(data, formula = votes ~ v57_density + c_65plus)
tab_model(m)
```

It seems that the BBB's support can be predicted by the population density (fewer votes in denser areas), but the age factor doesn't seem to influence it much.

**Exercise**: Let's change our focus and try to predict the VVD's support.
Modify the code above to predict the VVD's vote share using `c_65plus`, `v43_nl`, and `v132_income`. 

(Note: Be sure to rename the `bbb` data to `vvd` to avoid confusing R's code checking)

```{r lm-solution}
library(sjPlot)
vvd <- filter(results, party == "VVD")
data <- inner_join(vvd, demographics)
m <- lm(data, formula = votes ~ c_65plus + v43_nl + v132_income)
tab_model(m)
```

```{r lm-code-check}
grade_code()
```

What patterns emerge from this artwork?
Do the coefficients tell a story that surprises you?
Remember, just like interpreting art, understanding statistics requires a keen eye and an open mind.
