---
title: "Fun with R: A look at the Dutch voter"
output: 
  learnr::tutorial:
    progressive: true
runtime: shiny_prerendered
editor_options: 
  chunk_output_type: console
learnr_to_md_options:
  keep_output: false
---

```{r setup, include=FALSE, remove_for_md=TRUE}
knitr::opts_chunk$set(message = FALSE, warning = FALSE, fig.path = "img/")
library(learnr)
gradethis::gradethis_setup()
tutorial_options(
  exercise.reveal_solution = FALSE
)
```

<style>
.Info {
    background-color: rgb(204, 229, 255);
    border: 1px solid rgb(184, 218, 255);
    color: rgb(0, 64, 133);
    padding: 1em;
    margin: 1em;
}
.Info::before {
  font-weight: bold;
  content: "🛈 Information";
}

</style>

## CBS and Voting data

For this tutorial, we will use demographic data from CBS (the Dutch statistics bureau) and voting results from the election board. 
Both of these can be downloaded from the respective sources, but to make it easy we use cleaned up versions from this repository.

Note: As you can see below, `R` can read `csv` files (a common format that can also be imported or exported from, e.g., Excel or SPSS)
directly from the internet!

### Demographics data

After running the code below with the `► Run Code` button, you will see a *data frame* on the screen.
Data frames are the main data object in R. 

Generally, each row represents a unit of analysis or measurement (e.g., a case or respondent),
and each column represents a quantity of information about that row, e.g., a measured or computed variable. 

In this case, each row represents a Dutch municipality (*gemeente*), with the columns being a selection of demographic information collected from [CBS regional figures](https://opendata.cbs.nl/statline/#/CBS/nl/dataset/70072ned/table?ts=1691399765879):

```{r datademo, exercise=T}
library(tidyverse)
demographics <- read_csv("https://raw.githubusercontent.com/vanatteveldt/ccslearnr/master/data/dutch_demographics.csv")
demographics
```

```{r, include=FALSE}
# Somehow shinyapps.io does not always pick up all the dependencies
library(ggcorrplot)
library(xkcd)
library(ggthemes)
library(sjPlot)
```

Can you guess what the columns mean? Click 'continue' to get an explanation of each column. 

::: Info ::::::
Note that you cannot break anything in these example code boxes! 
Every time you hit the 'Run code' button, it will start over from scratch.
Moreover, if you feel that you have messed up the code, you can press the `↻ Start Over` button and the original code will run.
:::

### An explanation of the columns

| Variable        | Explanation                                                                  |
| --------------- | ---------------------------------------------------------------------------- |
| gm              | Unique code for each municipality                                            |
| gemeente        | Unique name of each municipality                                             |
| v01_pop         | Total population                                                             |
| v57_density     | Population Density                                                           |
| v43_nl          | Percentage of population without a migration history                         |
| v122_disposable | Average disposable income per household                                      |
| v132_income     | Average standardized income per household                                    |
| v142_wealth     | Median household wealth                                                      |
| v153_uitkering  | Percentage of population receiving social benefits (excluding state pension) |
| c_65plus        | Percentage of population older than 65                                       |

### Voting data

Similar to above, we can load voting data per municipality. 
The code below will load data for each party in each municipality in the 2023 provincial elections.

`R` also has powerful commands to clean up and restructure data. 
For example, the code below shows the results for a single municipality (Groningen), ordered by vote share.
What happens if you remove the `desc(..)` function from arrange, i.e., change `arrange(desc(votes))` into `arrange(votes)`?

**Exercise**: Can you change the code to show the results for the municipality of Amsterdam, ordered by party (starting from A)?

```{r datavotes, exercise=TRUE}
library(tidyverse)
results <- read_csv("https://raw.githubusercontent.com/vanatteveldt/ccslearnr/master/data/dutch_elections_2023ps.csv")
results |>
  filter(gemeente == "Groningen") |>
  arrange(desc(votes))
```

```{r datavotes-solution}
library(tidyverse)
results <- read_csv("https://raw.githubusercontent.com/vanatteveldt/ccslearnr/master/data/dutch_elections_2023ps.csv")
results |>
  filter(gemeente == "Amsterdam") |>
  arrange(party)
```

```{r datavotes-code-check}
grade_code()
```

```{r datavotes-hint-1}
# Use gemeente == "Amsterdam" within the filter() command
```

```{r datavotes-hint-2}
# Use gemeente == "Amsterdam" within the filter() command
# Replace the 'votes' in the arrange command by 'party', and remove the 'desc(.)' function
```

```{r datavotes-hint-3}
# Use gemeente == "Amsterdam" within the filter() command
# Replace the 'votes' in the arrange command by 'party', and remove the 'desc(.)' function
results |> 
  filter(gemeente == "Amsterdam") |> 
  arrange(party)
```

::: Info ::::::
The code box above is an exercise with a single correct solution. 
You need to change some code and click the `☑️Submit Answer` button to submit your answer.
Don't worry about making mistakes, you can try as often as you want and start over if needed.
Some exercises (like this one) also contain `💡 Hints` which you can click if you are stuck.
Also, if you submit an incorrect answer, `R` will do it's best to pinpoint where the problem is
(but of course it's only a computer, so it's not always very helpful...)
:::

## Simple visualizations

Besides the ability to read and clean up data, `R` has a very powerful visualization suite called `ggplot`. 
Although this can be hard to master, there are many useful resources to learn more and it can be very rewarding (and fun!) to make nice visualizations. 

For this section, we will use the same data as above:

```{r data, include=FALSE}
library(tidyverse)
results <- read_csv("https://raw.githubusercontent.com/vanatteveldt/ccslearnr/master/data/dutch_elections_2023ps.csv")
demographics <- read_csv("https://raw.githubusercontent.com/vanatteveldt/ccslearnr/master/data/dutch_demographics.csv")
```

```{r dataprep}
library(tidyverse)
results <- read_csv("https://raw.githubusercontent.com/vanatteveldt/ccslearnr/master/data/dutch_elections_2023ps.csv")
demographics <- read_csv("https://raw.githubusercontent.com/vanatteveldt/ccslearnr/master/data/dutch_demographics.csv")
```

### A first scatter plots

Let's make a first minimal scatter plot.
A scatter plot can be very useful to see the relation between variables, for example between population density and age:
do older people live mostly in cities or in more rural regions?

```{r scatter, exercise=T, exercise.setup='data'}
ggplot(
  data = demographics,
  mapping = aes(x = v57_density, y = c_65plus)
) +
  geom_point()
```

So, it turns out that older people mostly live in less dense municipalities. 
Have a quick look at the code above. The ggplot function takes two arguments, the data (`demographics`) 
and an *aesthetic mapping* (`aes(..)`). This mapping links columns in the the data frame (e.g. `v57_density`)
to graphical elements of the plot (e.g. the x position of the scatter points).
Finally, you add geometrical elements to the plot, in this case a `geom_point` or scatter point. 

What happens when you replace variables in the mapping with other variables, e.g. `v43_nl` or `v132_income`?
You can also try adding another aesthetic, for example `color = v43_nl` or `size = v01_pop`.

### A prettier plot

The plot above was informative, but not very pretty. 
The example below shows a much richer plot command.
It may look scary at first, but if you look at the commands they make more sense than you would think.
It also shows quite well how ggplot works: 
You first construct a basic plot consisting of data, aesthetics, and geoms. 
This generally produces a decent plot, as ggplot is quite good at guessing what good scales etc. are based on your data.
However, you can add many optional elements to the plot to change the appearance, from changing the color bars to repositioning the legend.
Take a look at the code below:

::: Info ::::::
Note: the code box below contains multiple *comments*. 
Every line starting with a `#` is called a comment, and is simply ignored by `R` so can be used to explain your code.
:::

```{r scatterpretty, exercise=T, exercise.setup='data'}
# Basic ggplot
#    - data (demographics)
#    - a mapping of density (y) against income (x), population size (dot size), and migration history (color),
#    - and a geom (in this case geom_point to create a scatter plot)
ggplot(
  data = demographics,
  mapping = aes(x = v57_density, y = v132_income, size = v01_pop, color = 100 - v43_nl)
) +
  geom_point(alpha = .7) +
  # All optional elements to change appearance
  # Add nice titles for the plot and axes
  ggtitle("Differences between rural and urban municipalities") +
  xlab("Population density") +
  ylab("% of people older than 65") +
  # Change the scale for migration history to go from light to dark blue:
  scale_color_gradient(low = "lightblue", high = "darkblue") +
  # Use the scale command to tweak the legend;
  scale_size(
    breaks = c(5000, 50000, 500000),
    labels = c("5.000", "50.000", "500.000")
  ) +
  # Use the guide command to specify titles and direction for the legend
  guides(
    size = guide_legend(
      title = "Population", title.position = "top",
      override.aes = list(color = "darkblue")
    ),
    color = guide_colorbar(
      title = "% with Migration History",
      title.position = "top",
      direction = "horizontal"
    )
  ) +
  # Use the 'classic' theme
  theme_classic() +
  # Tweak the theme to drop the background for legend keys
  theme(legend.key = element_blank())
```

In the code above, almost all lines are optional. What happens if you remove e.g. the theme or guide information?
And can you change the colors used in the color scale?

::: Info ::::::
In ggplot, many aspects of the appearance are controlled by *themes*, that specify things like fonts, colors, and grid lines.
Note that You can also use other themes, for example `theme_linedraw` or `theme_void`.
You can also install new themes. For example, the `ggthemes` packages contains a number of themes that mimic existing sources.
Try using `ggthemes::theme_economist` or `ggthemes::theme_tufte`.
:::

### Combining multiple data sources in a plot

Often, we need to combine data from multiple sources.
For example, it could be interesting to combine the election results in `results` with the demographic data.
Fortunately, in our example both sets have a common key variable (`gm`) on which to join them. 


::: Info ::::::
Combining data frames is very often needed while doing data analysis. In order to be able to combine data sets in R it is necessary both data sets have at least one variable that overlaps, in other words, a variable with the same name representing the same aspect of the data, e.g. an Id or a name of a municipality. In the latter case it is also needed that all these names are spelled exactly the same, since the computer cannot know that 's-Gravenhage is the same as Den Haag...
:::

So, let's see if we can plot the support for the farmer's party BBB against the population density:

```{r plot, exercise=T, exercise.setup='data'}
bbb <- filter(results, party == "BBB")
data <- inner_join(bbb, demographics)
ggplot(data = data, mapping = aes(x = log10(v57_density), y = votes, color = party, size = v01_pop)) +
  geom_point(alpha = .5) +
  xlab("Population density (log scale)") +
  ylab("Relative support for party") +
  scale_color_manual(values = c("BBB" = "green", "Anti-immigration" = "blue")) +
  scale_size(guide = "none") +
  ggtitle(
    "Support for BBB and anti-immigration parties per municipality",
    "(Dutch 2023 provincial elections; note: size of point relative to logged municipality population)"
  ) +
  xkcd::theme_xkcd() +
  theme(legend.position = "top", legend.title = element_blank())
```

## Plotting maps

Finally, to look at regional or national differences, it can be very insightful to plot variables on a map.

### Shape files

To do this, `R` first needs to know the *shape* of the various regions (such as the municipalities used above).
We can download a cleaned up Dutch 'shapefile' from the tutorial repository:

```{r loadshapes, include=FALSE, exercise.setup='data'}
library(sf)
if (!file.exists("sf_nl.rds")) {
  library(tidyverse)
  shapes <- read_rds("https://github.com/vanatteveldt/ccslearnr/raw/master/data/sf_nl.rds")
  write_rds(shapes, "sf_nl.rds")
} else {
  shapes <- read_rds("sf_nl.rds")
}
```


```{r datagis, exercise=TRUE}
library(tidyverse)
library(sf)
shapes <- read_rds("https://github.com/vanatteveldt/ccslearnr/raw/master/data/sf_nl.rds")
shapes
```

Now, we can plot this map using a regular ggplot command, using the `geom_sf` geometrical object, 
which can directly use the `geom` column from the shapefile. 
We can also specify the fill color of regions, for example using the province column:

```{r plotshape, exercise=TRUE, exercise.setup='loadshapes'}
ggplot(data = shapes) +
  geom_sf(aes(geometry = geom, fill = provincie))
```

### Combining shapes and demographic information

In the example above, we used only the spatial information to plot a map of Dutch municipalities.
Of course, it becomes much more interesting (and relevant!) if we can plot sociologically relevant information.

For this end, we need to `join` the spatial and demographic information, and then we can use other columns from the combined data set to color the regions. For example, the code below plots the population density per municipality:

```{r gisdensity, exercise=TRUE, exercise.setup='loadshapes'}
inner_join(shapes, demographics) |>
  ggplot() +
  geom_sf(aes(geometry = geom, fill = v57_density)) +
  scale_fill_gradient(low = "white", high = "red", guide = "none") +
  ggtitle("Population density per municipality") +
  theme_void()
```

We could compare this with the vote share of e.g. the (farmer's) party BBB:

```{r gisbbb, exercise=TRUE, exercise.setup='loadshapes'}
bbb <- results |> 
  filter(party == "BBB")
inner_join(shapes, bbb) |> 
  ggplot() +
  geom_sf(aes(geometry = geom, fill = votes)) +
  ggtitle("Support for BBB per municipality") +
  scale_fill_gradient(low = "white", high = "green", name = "% Support") +
  theme_void() +
  theme(legend.position = "bottom")
```

What does this tell us about the target constituency of the BBB party? 
Can you plot the same maps for other parties or variables? Can you change the color scheme?

## Simple statistics

Of course, `R` is also (and even at it's core) a statistical tool kit. 
In this sub section, we will look at some simple statistical results. 

### Correlation

First, let's see how the demographics are correlated. 
The command `cor` gives the bivariate correlations of all variables, quite useful for quick exploration:

```{r cor, exercise=TRUE, exercise.setup='data'}
vars <- select(demographics, v01_pop:c_65plus)
cor(vars, use = "pairwise")
```

We can also plot this using the `ggcorrplot` package:

```{r corplot, exercise=TRUE, exercise.setup='data'}
library(ggcorrplot)
vars <- select(demographics, v01_pop:c_65plus)
correlations <- cor(vars, use = "pairwise")
ggcorrplot(correlations, method = "circle")
```


### Regression

Let's see if we can predict/explain support for BBB from the demographic information.
For this, we combine the demographic and vote data as above, 
and use the `lm` function to compute the linear regression model.
Finally, we use the `tab_model` function from the `sjPlot` package to create a nice looking regression table:

```{r lm, exercise=TRUE, exercise.setup='data'}
library(sjPlot)
bbb <- filter(results, party == "BBB")
data <- inner_join(bbb, demographics)
m <- lm(data, formula = votes ~ v57_density + c_65plus)
tab_model(m)
```

```{r lm-solution}
library(sjPlot)
vvd <- filter(results, party == "VVD")
data <- inner_join(vvd, demographics)
m <- lm(data, formula = votes ~ c_65plus + v43_nl + v132_income)
tab_model(m)
```

```{r lm-code-check}
grade_code()
```

So, BBB's vote share can be predicted from population density (the denser the municipality, the fewer votes),
but not by the percentage of people over 65.

**Exercise**: Change the code above predict the VVD's vote share from `c_65plus`, `v43_nl`, and `v132_income`. 
How do you interpret this? Do the coefficients surprise you?

(Note: Be sure to rename the `bbb` data to `vvd` to avoid confusing R's code checking)
