---
title: "Assignment 1 template"
output: html_document
date: "2023-09-01"
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning=FALSE)
library(printr)
```

# Introduction

The goal of this assignment is to use the skills from the tutorial to run some simple analyses.
Substantively, we want to see which parties do better in larger municipalities. 

## Data

The code below creates a dataset containing voting results and municipality size.
Don't worry if you don't fully understand this code, the `join` commands will be explained
in a later tutorial. 

```{r data}
library(tidyverse)
results <- read_csv('https://raw.githubusercontent.com/vanatteveldt/ccslearnr/master/data/dutch_elections_2023ps.csv')
demographics <- read_csv('https://raw.githubusercontent.com/vanatteveldt/ccslearnr/master/data/dutch_demographics.csv')
data <- inner_join(results, demographics) |>
  select(gemeente, population=v01_pop, party, votes)
head(data)
```

## Filtering

To start, choose a political party, e.g. `GROENLINKS` (greens). 
**Filter** the results data frame to only contain that party, 
and store it as a new data frame called `subset`. 
Next, **arrange** the results to sort by votes (descending), and show the first rows of data:

Remove the comment character (`#`) from the lines below and fill in the blanks:

```{r filter}
# subset = filter(data, ____)
# subset = arrange(subset, ____)
# head(subset)
```

Interpretation: What municipalities is that party most succesful in? Does that make sense?
(For example, you would expect the greens to do well in cities, especially cities with many students)

*(Replace this sentence by your interpretation)*

## Scatter plot

Now, let's see to what extend the party is more succesful in larger or smaller municipalities.

First, let's first create a scatter plot to see to what extent their voting seems related.
In the code below, remove the `#` sign and fill in the blanks so the population is on the x-axis, and the votes are on the y-axis:

```{r scatter}
# ggplot(subset, aes(x=___, y=___)) + geom_point()x
```

Interpretation: How do you interpret the scatter plot? Does there seem to be a relation between population size and votes?

Bonus: If you want, you can add `+ scale_x_log10()` to the plot to set the population on a logarithmic scale. Does that make it easier to interpret?

Bonus2: If you want, you can add `+ ggtitle("my title") + ylab("% of votes") + xlab("Population") + theme_classic()` to the plot to add labels and change the theme.

## Correlation

After the explorative analysis above, let's do a simple statisitcal analysis to see if there is a significant correlation between vote share and population size.

Hint: the function `cor.test` is not part of the tidyverse, and expects you to select the columns from the data frame. 
You can do this using e.g. `subset$votes` to select the column called votes:

```{r correlation}
# cor.test(subset$____, subset$____)
```

## Bonus: Map

As a bonus exercise, can you adapt the code from the 'Fun with R' tutorial to create a map of vote shares?

## Submitting this assignment

To submit the assignment, **knit** it to a word document, download it, and submit it on canvas. 
Alternatively, you can print the html report as a pdf and submit that. 
