---
title: "Cooking with Data: A Gourmet Look at the Dutch Voter"
output: 
  learnr::tutorial:
    progressive: true
runtime: shiny_prerendered
editor_options: 
  chunk_output_type: console
learnr_to_md_options:
  keep_output: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(message = FALSE, warning = FALSE, fig.path = "img/")
library(learnr)
gradethis::gradethis_setup()
tutorial_options(
  exercise.reveal_solution = FALSE
)
```

<style>
.Info {
    background-color: rgb(204, 229, 255);
    border: 1px solid rgb(184, 218, 255);
    color: rgb(0, 64, 133);
    padding: 1em;
    margin: 1em;
}
.Info::before {
  font-weight: bold;
  content: "🍽️ Chef's Tip";
}

</style>

## Ingredients

For our gourmet dish, we will be using:

- Demographic data from CBS (the Dutch statistics bureau)
- Voting results from the election board

Both of these ingredients can be sourced fresh from their respective sources, but for convenience, we have pre-cleaned versions available.

::: Info ::::::
Remember, `R` can fetch `csv` files directly from the internet, making it easier to gather fresh ingredients!
:::

### Preparing the Demographics

After executing the code below, you will have a beautifully prepared *data frame* ready for cooking.
Data frames are like the favorite cooking pan of `R` as every dish is a bit more tasty when it is prepared in the well-used iron.
Here, the data has already been carefully placed so each row is a unique flavor (municipalities/*gemeente*) and each column adds a different texture (variable).

Our dish will feature Dutch municipalities (*gemeente*), with various demographic seasonings from [CBS regional figures](https://opendata.cbs.nl/statline/#/CBS/nl/dataset/70072ned/table?ts=1691399765879):

```{r datademo, exercise=T}
library(tidyverse)
demographics <- read_csv("https://raw.githubusercontent.com/vanatteveldt/ccslearnr/master/data/dutch_demographics.csv")
demographics
```

```{r, include=FALSE}
# Somehow shinyapps.io does not always pick up all the dependencies
library(ggcorrplot)
library(xkcd)
library(sjPlot)
```

Curious about the flavors in our dish?
Try to guess first what each column is.
Click `continue` for a breakdown of each seasoning.

::: Info ::::::
Remember, in our kitchen, you can't ruin the dish!
Every time you execute the code, it's like starting with fresh ingredients.
If you feel you've added too much salt, just hit the `↻ Start Over` button and begin anew.
:::

### Seasoning Breakdown

| Ingredient      | Flavor Profile                                                               |
| --------------- | ---------------------------------------------------------------------------- |
| gm              | Unique code for each municipality                                            |
| gemeente        | Unique name of each municipality                                             |
| v01_pop         | Total population                                                             |
| v57_density     | Population Density                                                           |
| v43_nl          | Percentage of population without a migration history                         |
| v122_disposable | Average disposable income per household                                      |
| v132_income     | Average standardized income per household                                    |
| v142_wealth     | Median household wealth                                                      |
| v153_uitkering  | Percentage of population receiving social benefits (excluding state pension) |
| c_65plus        | Percentage of population older than 65                                       |

### Preparing the Voting Data

Just like our demographics, we need to prepare the voting data.
This will add a rich flavor to our dish, representing each party in each municipality from the 2023 provincial elections.
For example, the code below shows the results for a single municipality (Groningen), ordered by vote share.
What happens if you remove the `desc(..)` function from arrange (change `arrange(desc(votes))` into `arrange(votes)`)?

**Chef's Challenge**: After preparing the data for Groningen, can you adjust the recipe to showcase the flavors of Amsterdam, ordered by party (from A to Z)?

```{r datavotes, exercise=TRUE}
library(tidyverse)
results <- read_csv("https://raw.githubusercontent.com/vanatteveldt/ccslearnr/master/data/dutch_elections_2023ps.csv")
results |>
  filter(gemeente == "Groningen") |>
  arrange(desc(votes))
```

```{r datavotes-solution}
library(tidyverse)
results <- read_csv("https://raw.githubusercontent.com/vanatteveldt/ccslearnr/master/data/dutch_elections_2023ps.csv")
results |>
  filter(gemeente == "Amsterdam") |>
  arrange(party)
```

```{r datavotes-code-check}
grade_code()
```

```{r datavotes-hint-1}
# To bring out the flavors of Amsterdam, use gemeente == "Amsterdam" in the filter() command
```

```{r datavotes-hint-2}
# To bring out the flavors of Amsterdam, use gemeente == "Amsterdam" in the filter() command
# Adjust the seasoning by replacing 'votes' in the arrange command with 'party', and remove the 'desc(.)' function
```

```{r datavotes-hint-3}
# To bring out the flavors of Amsterdam, use gemeente == "Amsterdam" in the filter() command
# Adjust the seasoning by replacing 'votes' in the arrange command with 'party', and remove the 'desc(.)' function
results |> 
  filter(gemeente == "Amsterdam") |> 
  arrange(party)
```

::: Info ::::::
Remember, cooking is an art!
The code box above is like a cooking challenge.
Adjust the recipe and click the `☑️Submit Answer` button to see if you've mastered the dish.
Don't worry about mistakes; every chef has their unique style.
If you're stuck, click on the `💡 Hints` for some culinary guidance.
:::

## Plating Your Data: Simple Visualizations

In our data kitchen, we have a special tool called `ggplot` that allows us to plate our data in the most visually appealing way!
Mastering `ggplot` is like mastering the art of plating in gourmet cooking.
It might seem challenging at first, but with practice, you can create visual masterpieces.

For this part, we'll be using our previously prepared ingredients:

```{r data, include=FALSE}
library(tidyverse)
results <- read_csv("https://raw.githubusercontent.com/vanatteveldt/ccslearnr/master/data/dutch_elections_2023ps.csv")
demographics <- read_csv("https://raw.githubusercontent.com/vanatteveldt/ccslearnr/master/data/dutch_demographics.csv")
```

```{r dataprep}
library(tidyverse)
results <- read_csv("https://raw.githubusercontent.com/vanatteveldt/ccslearnr/master/data/dutch_elections_2023ps.csv")
demographics <- read_csv("https://raw.githubusercontent.com/vanatteveldt/ccslearnr/master/data/dutch_demographics.csv")
```

### Crafting Your First Scatter Plot

Imagine a scatter plot as a plate sprinkled with ingredients.
Now, each ingredient you place on this plate symbolizes a row of data from our dataset.
The position where you place each ingredient on the plate is determined by the values of specific variables for that particular row of data.
Let's create our first scatter plot to see the relationship between population density and age.
Are our older ingredients mostly found in urban or rural areas?

```{r scatter, exercise=T, exercise.setup='data'}
ggplot(
  data = demographics,
  mapping = aes(x = v57_density, y = c_65plus)
) +
  geom_point()
```

It seems older restaurant customers are mostly found in less dense parts of the country.
Dive into the recipe above. 
The `ggplot` function uses our ingredients (`data = demographics`) and a *presentation style* (`aes(..)`).
This style links our ingredients to the visual elements of our plate (`v57_density` dictates the placement on the x, `c_65plus` on the y axis).

**Chef's Challenge**: Can you sprinkle other ingredients on our plate, like `v43_nl` or `v132_income`?
 Try adding more flair with `color = v43_nl` or `size = v01_pop`.

### Elevating Your Presentation

Our initial plate was delightful, but let's elevate it!
The recipe below might seem intricate, but each step adds a unique touch to our presentation.
It also shows quite well how ggplot works: 
You first construct a basic plot consisting of data, aesthetics, and geoms. 
This generally produces a decent plot, as ggplot is quite good at guessing what how the data can be best presented.
However, like changing a proven recipe, you can add many optional elements to the plot to change the appearance, from changing the color bars to repositioning the legend.
Just the way you like to eat (I mean look at) it.
Dive in and see how `ggplot` works:

::: Info ::::::
Remember, every line starting with a `#` is a chef's note, guiding you through the recipe.
`R` does not execute them, but treats everything right of the `#` like it wasn't there.
:::

```{r scatterpretty, exercise=T, exercise.setup='data'}
# Basic ggplot
#    - data (demographics)
#    - a mapping of density (y) against income (x), population size (dot size), and migration history (color),
#    - and a geom (in this case geom_point to create a scatter plot)
ggplot(
  data = demographics,
  mapping = aes(x = v57_density, y = v132_income, size = v01_pop, color = 100 - v43_nl)
) +
  geom_point(alpha = .7) +
  # All optional elements to change appearance
  # Add nice titles for the plot and axes
  ggtitle("Differences between rural and urban municipalities") +
  xlab("Population density") +
  ylab("% of people older than 65") +
  # Change the scale for migration history to go from light to dark blue:
  scale_color_gradient(low = "lightblue", high = "darkblue") +
  # Use the scale command to tweak the legend;
  scale_size(
    breaks = c(5000, 50000, 500000),
    labels = c("5.000", "50.000", "500.000")
  ) +
  # Use the guide command to specify titles and direction for the legend
  guides(
    size = guide_legend(
      title = "Population", title.position = "top",
      override.aes = list(color = "darkblue")
    ),
    color = guide_colorbar(
      title = "% with Migration History",
      title.position = "top",
      direction = "horizontal"
    )
  ) +
  # Use the 'classic' theme
  theme_classic() +
  # Tweak the theme to drop the background for legend keys
  theme(legend.key = element_blank())
```

In the recipe above, each step adds a unique touch.
 What happens if you adjust the presentation style or colors?

::: Info ::::::
In `ggplot`, presentation styles are controlled by *themes*. You can experiment with different themes like `theme_linedraw` or `theme_void`.
For a gourmet touch, try using `ggthemes::theme_economist` or `ggthemes::theme_tufte`, which are additional styles from the package `ggthemes`.
:::

### Layering Flavors: Combining Multiple Data Sources

In gourmet cooking, combining ingredients from different kitchens can lead to a symphony of flavors, like in a fusion restaurant.
Similarly, in data visualization, combining data from multiple sources can provide richer insights.

Imagine you want to understand the preference for a particular ingredient (the farmer's party BBB) across different regions, and how it relates to the base flavor (population density).
We first `join` them, which works automatically because they both have the `gm` column.
Here's how you can visualize this:

```{r plot, exercise=T, exercise.setup='data'}
bbb <- filter(results, party == "BBB")
data <- inner_join(bbb, demographics)
ggplot(data = data, mapping = aes(x = log10(v57_density), y = votes, color = party, size = v01_pop)) +
  geom_point(alpha = .5) +
  xlab("Population density (log scale)") +
  ylab("Preference for Ingredient (BBB)") +
  scale_color_manual(values = c("BBB" = "green", "Anti-immigration" = "blue")) +
  scale_size(guide = "none") +
  ggtitle(
    "Preference for BBB and Anti-immigration Ingredients Across Regions",
    "(Dutch 2023 provincial elections; note: size of point relative to logged region population)"
  ) +
  xkcd::theme_xkcd() +
  theme(legend.position = "top", legend.title = element_blank())
```

## Plating on Maps: Visualizing Regional Differences

To truly appreciate regional differences in flavor preferences, we need to plate our data on a map.

### Preparing the Map: Shape Files

Before we can plate our data, we need a map or a 'shape' of the regions.
Think of this as the outline of our plate:


```{r loadshapes, include=FALSE, exercise.setup='data'}
library(sf)
if (!file.exists("sf_nl.rds")) {
  library(tidyverse)
  shapes <- read_rds("https://github.com/vanatteveldt/ccslearnr/raw/master/data/sf_nl.rds")
  write_rds(shapes, "sf_nl.rds")
} else {
  shapes <- read_rds("sf_nl.rds")
}
```


```{r datagis, exercise=TRUE}
library(tidyverse)
library(sf)
shapes <- read_rds("https://github.com/vanatteveldt/ccslearnr/raw/master/data/sf_nl.rds")
shapes
```

With our plate ready, let's start plating our ingredients:

```{r plotshape, exercise=TRUE, exercise.setup='loadshapes'}
ggplot(data = shapes) +
  geom_sf(aes(geometry = geom, fill = provincie))
```

`geom_sf` knows how to work with the ingredients in `shapes`.
We also used the `provincie` to give our map that extra tasty color.

### Layering with Demographic Information

Now, let's add some flavor to our map by layering it with demographic information.
This will help us understand the base flavors of each region.

We first use `join` to combine the data evenly in our pan, then we simmer it nicely to bring out the population density flavour:

```{r gisdensity, exercise=TRUE, exercise.setup='loadshapes'}
inner_join(shapes, demographics) |>
  ggplot() +
  geom_sf(aes(geometry = geom, fill = v57_density)) +
  scale_fill_gradient(low = "white", high = "red", guide = "none") +
  ggtitle("Base Flavors (Population Density) per Region") +
  theme_void()
```

To understand the preference for our special ingredient (BBB), let's visualize its popularity across regions:

```{r gisbbb, exercise=TRUE, exercise.setup='loadshapes'}
bbb <- results |> 
  filter(party == "BBB")
inner_join(shapes, bbb) |> 
  ggplot() +
  geom_sf(aes(geometry = geom, fill = votes)) +
  ggtitle("Popularity of Ingredient (BBB) per Region") +
  scale_fill_gradient(low = "white", high = "green", name = "% Preference") +
  theme_void() +
  theme(legend.position = "bottom")
```

**Chef's Challenge**: What insights can you gather about the preference for the BBB ingredient? Can you plate similar maps for other ingredients or flavors? How about experimenting with different color palettes?

Let's wrap up our culinary data journey by diving into the essence of data analysis: statistics. Think of this as the taste testing phase, where we evaluate the flavors and textures of our dish to ensure it's just right.

## Taste Testing with Simple Statistics

In the world of data, `R` is our taste testing toolkit. It allows us to evaluate the relationships and patterns in our data, ensuring we have a well-balanced dish.

### Evaluating Flavors: Correlation

Correlation is like tasting two ingredients together to see how well they complement each other. Let's see how our demographic flavors correlate:

```{r cor, exercise=TRUE, exercise.setup='data'}
vars <- select(demographics, v01_pop:c_65plus)
cor(vars, use = "pairwise")
```

To visualize these correlations, we can use the `ggcorrplot` package, which is like presenting a flavor wheel of our ingredients:

```{r corplot, exercise=TRUE, exercise.setup='data'}
library(ggcorrplot)
vars <- select(demographics, v01_pop:c_65plus)
correlations <- cor(vars, use = "pairwise")
ggcorrplot(correlations, method = "circle")
```


### Crafting the Perfect Blend: Regression

Regression is like adjusting the quantities of our ingredients to achieve the perfect balance.
Let's see if we can predict the preference for our BBB ingredient using the demographic flavors:

```{r lm, exercise=TRUE, exercise.setup='data'}
library(sjPlot)
bbb <- filter(results, party == "BBB")
data <- inner_join(bbb, demographics)
m <- lm(data, formula = votes ~ v57_density + c_65plus)
tab_model(m)
```

From our taste test, we see that the preference for BBB is influenced by population density, but not by the percentage of older flavors (people over 65).

**Chef's Challenge**: Can you adjust the recipe to predict the preference for the VVD ingredient using `c_65plus`, `v43_nl`, and `v132_income`? How do these flavors influence the preference? Are you surprised by the blend?

```{r lm-solution}
library(sjPlot)
vvd <- filter(results, party == "VVD")
data <- inner_join(vvd, demographics)
m <- lm(data, formula = votes ~ c_65plus + v43_nl + v132_income)
tab_model(m)
```

```{r lm-code-check}
grade_code()
```

(Note: Remember to rename the `bbb` data to `vvd` to ensure our taste testing toolkit (R) understands your adjustments.)

Bon Appétit! With your statistical insights, you're now ready to serve a well-balanced and flavorful data dish!

