---
title: "Recoding data in R"
output: 
  learnr::tutorial:
    progressive: true
runtime: shiny_prerendered
editor_options: 
  chunk_output_type: console
learnr_to_md_options:
  keep_output: false
---

```{r setup, include=FALSE, remove_for_md=TRUE}
knitr::opts_chunk$set(message = FALSE, warning = FALSE, fig.path = "img/")
library(learnr)
gradethis::gradethis_setup()
```

## Introduction: Recoding variables

Quite often, the data you have is not quite the data you'd like to have. 
For example, dates might be in the wrong format, variables might be coded the wrong way,
it might contain typos or other mistakes, etc.

In this tutorial, you will learn about a number of common recoding steps that you can use to solve such issues. 

## Recoding values

Often, our data is categorized or coded differently from how we would like to analyse it. 
Fixing this is generally called 'recoding', and in this section you will learn three tidyverse functions for achieving this. 

### Why Recoding?

For example, for the 2021 elections we received data from a panel survey company that looked like this:

```{r}
library(tidyverse)
votes = read_csv("https://raw.githubusercontent.com/vanatteveldt/ccslearnr/master/data/vu_election_2021.csv")
head(votes)
```

As you can see, this lists individual respondents with their age, education, etc, and their voting behaviour in 2017 and 2021.
Note that the 2021 data contains many missing values as not everyone who started the panel survey finished it.

Suppose we want to make an overview of e.g. TV use by age and education (like in [our analysis of the first wave](https://tk2021.vupolcom.nl/reports/wave0/#media-use-by-age-and-education)).
This means we would need to categorize age into age groups, and change the categorization of the education column.



### Creating two categories: `if_else`

Om the simplest case, you might want to recode into two categories. 
For example, is someone older than 65, did someone vote PVV or not, is someone of Dutch origin or not?

In such cases, `if_else` is the easiest way to recode values. This function takes 4 arguments: `if_else(condition, true, false, missing)`, where `missing` is optional. 
The condition is a comparison similar to when you use `filter`, e.g. `age >= 65`. For each row, if this condition is True (i.e. the respondent is indeed older than 65), it will assign the `true` value, otherwise it will assign the `false` value. 
If `age` is missing (NA), it will assign the `missing` value. If you don't specify a value for missing cases, they will simply assign `NA`.


```{r readvotes, remove_for_md=TRUE}
library(tidyverse)
votes = read_csv("https://raw.githubusercontent.com/vanatteveldt/ccslearnr/master/data/vu_election_2021.csv")
head(votes)
```
```{r ifelse, exercise=TRUE, exercise.setup='readvotes'}
votes |> 
  mutate(is_65plus = if_else(age >= 65, "65+", "Not 65+")) |>
  select(iisID, age, is_65plus)
```
```{r ifelse-solution}
votes |> 
  mutate(is_65plus = if_else(age >= 65, "65+", "Not 65+"),
         origin = if_else(ethnicity == 'Dutch origin', 'Dutch','other')) |>
  select(iisID, age, is_65plus, origin)
```
```{r ifelse-code-check}
grade_code()
```

Exercise: In addition to the is_65plus column above, can you create a column 'origin' with two values, 'Dutch' and 'other'?


::: Info ::::::
Note that base R also contains a very similar function, `ifelse` (without the underscore).
Although they will do the same in most cases. we advise using the tidyverse fucntion `if_else` for two reasons.
First, it has explicit support for dealing with missing values, which can often be quite useful.
Second, there are some cases in which `ifelse` gives an undesired result, especially when dealing with date values.
:::

### Creating multiple categories: `case_match`

If we want to recode existing values into more than one category, we can use `case_match(column, recode)`.
The recode arguments match values from the column, and specify which new category to assign.
They do this by (ab)using the **formula** notation that you might know from statistical models.
In case_match, each recode argument looks like `value(s) ~ category`. 
If the column matches the value(s) in the argument, it's assigned that category.
If if matched none of the arguments, if will be assigned `NA` unless you specify a default value with `.default=value`.

This sounds very complicated, so let's look at an example.
The following code recodes the education into three categories:

```{r case_match, exercise=TRUE, exercise.setup='readvotes'}
votes |> mutate(edu_level=case_match(
  education,
  c("VMBO-T", "VMBO, MBO 1", "Elementary") ~ "Low",
  c("MBO 2-4", "HAVO/VWO") ~ "Middle",
  c("Bachelor", "Master") ~ "High"
)) |> select(iisID, education, edu_level)
```

### Checking recoding results with `summarize`

When recoding, it is often a good idea to check how the old and new values correspond. 
We can group by on the old and new value, and use `summarize` to count how often each case arises.
For example, let's consider the following 

```{r jobs, exercise=TRUE, exercise.setup='readvotes'}
votes |> mutate(job_category=case_match(
  job,
  c("Full-time employed", "Part-time employed") ~ "Employed",
  .default="Other"
)) |> group_by(job, job_category) |>
  summarize(n=n()) |> 
  arrange(job_category, desc(n))
```
```{r jobs-solution}
votes |> mutate(job_category=case_match(
  job,
  c("Full-time employed", "Part-time employed", "Entrepreneur") ~ "Employed",
  c("Unemployed and not searching for a job or incapacitated", "Unemployed and searching for a job") ~ "Unemployed",
  .default="Other"
)) |> group_by(job, job_category) |>
  summarize(n=n()) |> 
  arrange(job_category, desc(n))
```
```{r jobs-code-check}
grade_code()
```

**Exercise**: Can you update the code above to also include 'enterpreneur' in the employed category,
and to create a category for unemployed that contains both unemployed conditions?

### Recoding any value: `case_when`

The final function for recoding, `case_when`, is quite similar to the case_match function you played with earlier. 
You also specify conditions and values using the `condition ~ value` format. 
The big difference is that where `case_match` used simple values or value lists as conditions,
with `case_when` you specify the condition as a comparison.

Again, let's look at an example first:

```{r age, exercise=TRUE, exercise.setup='readvotes'}
votes |> mutate(age_category=case_when(
  age <= 30 ~ "Younger than 30",
  age < 65 ~ "Between 31 and 65",
  age >= 65 ~ "65+"
)) |> select(iisID, age, age_category)
```

```{r age-solution}
votes |> mutate(age_category=case_when(
  age <= 30 ~ "Younger than 30",
  age < 45 ~ "Between 31 and 45",
  age < 65 ~ "Between 46 and 65",
  age >= 65 ~ "65+"
)) |> select(iisID, age, age_category)
```
```{r age-code-check}
grade_code()
```
**Exercise**: Can you split the 31 to 65 category into two categories, "Between 31 and 45" and "Between 46 and 65"?

An important thing to note in the code above is that you don't need to specify e.g. `age > 30 & age < 65` for that category.
The reason for that is that the conditions are evaluated from top to bottom. 
Thus, anyone younger than 30 will already be dealt with by the first category, so the second condition is never checked. 
As with case_match, you can set a `.default` value for any case not matched by the conditions, e.g. because of missing values. 

While this last function is a bit more verbose than `case_match`, it is incredibly useful in two cases.
First, like the example above, for categorizing values into ranges it is better to specify the range with a comparison than as a list of all ages belonging to the category. 
Secondly, you can use aribitrary functions from different columns. 
For example, we could use `str_detect(job, 'Unemployed') ~ "unemployed"` to find all cases that include the word 'Unemployed', rather than having to specify them all. 
We could also combine values from different columns, e.g. with something like `age < 30 & str_detect(job, 'Unemployed') ~ "Unemployed youth"` 

The advice is generally to use the simplest function possible. If `if_else` does the job, it's the easiest function to use. However, learning how to use `case_match` and `case_when` gives you acess to two very flexible recoding functions.

## Dates and date formats

We are often interested in how things change over time, for which we need to work with date and time data.
In this tutorial, we go over three of the most common use cases: converting text to date,
formatting dates as text, and computing the year, month or week of a date (so we can group by e.g. year).

These examples all use functions from tidyverse's `lubridate` package,
which is automatically loaded when you load tidyverse. 
See also the [lubridate page](https://lubridate.tidyverse.org/), which also links to the R4DS chapter and cheat sheet.

Note: `lubridate` is part of the `tidyverse`, and should already be installed in your system.
However, it is not loaded automatically when you `library(tidyverse)`. 
So, whenever you use the lubridate functions, explicitly activate the package with 

```{r}
library(lubridate)
```

### Converting text to dates

Many data sets contain a date or timestamp column, but often these are not directly recognized as dates.
For example, they might be in German (4.11.2016), American (11/4/2016) or textual foramt (4 November 2016).

The good news is that the `parse_date_time` functions make this quite easy.
The bad news is that you still need to tell the computer the **format** used:
Is `11/4/2016` is a date in April or in November?

For example, the call below parses that date assuming it's American notation:

```{r parse_date, exercise=TRUE}
library(lubridate)
date <- parse_date_time("11/4/2016", "dmy")
# date
class(date)
```

The code above converts the **text** "11/4/2016" in a Date object which is represented as "2016-11-04". 
It then checks the data type (or `class`) of the object, which is `POSIXct`, which for R means a date plus time.
Since R now knows that it's a date value, you can e.g. use +1 to add one day to the value.
Try adding or subtracting a large number of days, and you will see that R understand months, years, and even leap years. 

### Format strings

In the `parse_date_time` function above, a crucial role is played by the second (`order`) argument.
This describes how to read the date. 
In that case, we are looking for a month number (`m`), day number (`d`), and a year number (`y`).

For a full list of possible codes in this argument (like `m`), see the [parse_date_time documentation](https://lubridate.tidyverse.org/reference/parse_date_time.html). The table below lists some of the most frequently used codes:

| Code | Meaning |
| ----- | ------- |
| y | The year, i.e. 2023 or 23 |
| m | The month i.e. 4 or April |
| d | The day of the month, i.e. 11 for April 11 |
| H | The hour, i.e. 0-24 |
| M | The minute, i.e. 0:59 |
| S | The second, i.e. 0:59 |


Can you parse all of the dates below?

```{r parsedate, exercise=TRUE, eval=FALSE}
library(lubridate)
parse_date_time("4.11.2016", ____)
parse_date_time("04-11-16", ____)
parse_date_time("4 November 2016", ____)
parse_date_time("November 4th, 2016", ____)
parse_date_time("04-11-2016 11:23", ____)
```
```{r parsedate-solution, exercise=TRUE}
library(lubridate)
parse_date_time("4.11.2016", "dmy")
parse_date_time("04-11-16", "dmy")
parse_date_time("4 November 2016", "dmy")
parse_date_time("November 4th, 2016", "mdy")
parse_date_time("04-11-2016 11:23", "dmy HM")
```
```{r parsedate-code-check}
grade_code()
```


### Exercise

In 2015, the data science blog 538 had an [item about the US satirical TV show *The Daily Show*](https://fivethirtyeight.com/features/every-guest-jon-stewart-ever-had-on-the-daily-show/), for which they kindly made their data available. 
As this is shared in a CSV format, R inteprets the `Show` column (which has the date of the show) as a textual (string) variable.
Can you convert it into a Date column?

```{r daily, exercise=TRUE, eval=FALSE}
library(tidyverse)
library(lubridate)
daily = read_csv("https://raw.githubusercontent.com/fivethirtyeight/data/master/daily-show-guests/daily_show_guests.csv")
# daily <- daily |> mutate(date = parse_date_time(Show, ____))
head(daily)
```

```{r daily-solution, exercise=TRUE}
library(tidyverse)
library(lubridate)
daily = read_csv("https://raw.githubusercontent.com/fivethirtyeight/data/master/daily-show-guests/daily_show_guests.csv")
daily <- daily |> mutate(date = parse_date_time(Show, "mdy"))
head(daily)
```
```{r daily-code-check}
grade_code()
```

### Weeks, Months, and years 

Suppose we would want to analyse changes in the Daily show (or political results, or income and poverty ) over time.
Often, our date (like above) is at the daily level, but for a model or graph we generally want to group it at the monthly or weekly level.
To do this, we can use the `floor_date` function, which rounds a date down to the first day of the month, week, or quarter it's in. 

The syntax for floor_date is floor_date(date, unit), where unit can be e.g. "week", "month", or "quarter".

Can you create a plot of the number of guests per category for each quarter for the daily show?

Tip: If you select part of the code and press Run Code, only the selected part is execute.
This allows you to e.g. first only run the first step and work on the second step once that works. 

```{r dailyguest, exercise=TRUE, eval=FALSE}
library(tidyverse)
library(lubridate)
daily = read_csv("https://raw.githubusercontent.com/fivethirtyeight/data/master/daily-show-guests/daily_show_guests.csv")
daily <- daily |> 
  mutate(Date = parse_date_time(Show, "mdy")) |>
  mutate(Quarter = ____)  # Compute the Quarter from the Date value
perquarter <- daily |> group_by(Quarter, Group) |> 
  summarize(n=n()) |>
  mutate(Percentage=____)  # Compute the percentage from the count (hint: and the group total, i.e. the sum(n))
ggplot(perquarter, aes(____)) + # Set the right aesthetics
  geom_line() 
```

```{r dailyguest-solution}
library(tidyverse)
library(lubridate)
daily = read_csv("https://raw.githubusercontent.com/fivethirtyeight/data/master/daily-show-guests/daily_show_guests.csv")
daily <- daily |> 
  mutate(Date = parse_date_time(Show, "mdy")) |>
  mutate(Quarter = floor_date(Date, "quarter"))
perquarter <- daily |> group_by(Quarter, Group) |> 
  summarize(n=n()) |>
  mutate(Percentage=n/sum(n))
ggplot(perquarter, aes(x=Quarter, y=Percentage, color=Group)) +
  geom_line() 
```
```{r dailyguest-code-check}
grade_code()
```


As an extra challenge (which is not included in the 'official answer' above), 
can you use the recode functions from the start of the tutorial to clean up the guest categories,
i.e. combine media and Media, and create categories for all science or politics related guests?
