---
title: 'Visualizing data with ggplot'
author: "Wouter van Atteveldt"
output: 
  learnr::tutorial:
    progressive: true
runtime: shiny_prerendered
editor_options: 
  chunk_output_type: console
learnr_to_md_options:
  keep_output: false
---

```{r setup, include=FALSE, remove_for_md=TRUE}
knitr::opts_chunk$set(message=FALSE, warning=FALSE, fig.path = "img/")
library(learnr)
gradethis::gradethis_setup()
tutorial_options(
  exercise.reveal_solution = FALSE
)
```
<style>
.Info {
    background-color: rgb(204, 229, 255);
    border: 1px solid rgb(184, 218, 255);
    color: rgb(0, 64, 133);
    padding: 1em;
    margin: 1em;
}
.Info::before {
  font-weight: bold;
  content: "🛈 Information";
}
</style>

## Introduction

Creating visualizations is one of the most useful and powerful ways to analyse and data
and tell a convincing story with these data. 

In `ggplot`, R Tidyverse has a very powerful and well-designed visualization library.
It can be used to make almost all kinds of graphs, 
from simple line and scatter plots to colorful annotated bubble plots. 

For an an overview of possibilities and more explanation, you can check out the free online resources below:

- [The R Graph gallery](https://r-graph-gallery.com/ggplot2-package.html) - contains many examples of beautiful plots, all with R code included
- [From data to viz](https://www.data-to-viz.com/) - an interactive site that showcases good plots based on your data, giving not just the R code but also best practices, common mistakes, and alternatives for each plot. 
- [The ggplot cheat sheet](https://github.com/rstudio/cheatsheets/blob/main/data-visualization.pdf) with a nice overview of ggplot functions and options
- [Data Visualization](https://socviz.co/) by Kieran Healy - a completely free online book that discusses the principles of data visualization and the  best ways to visualize various types of data, also including R code for all examples.


### GGplot basics

To illustrate how ggplot works, let's take a look at a simple built-in data set of car tests:

```{r}
library(tidyverse)
data(mtcars)
mtcars
```

```{r mtcars}
library(tidyverse)
data(mtcars)
```

We can visualize the relation between engine power and fuel economy:

```{r basicplot, exercise=TRUE, exercise.setup='mtcars'}
ggplot(mtcars, 
       aes(x=hp, y=mpg)) + 
  geom_point()
```


This ggplot call is typical for how ggplot works,
and showcases the *grammar of graphics*. Essentially, every graph contains these three basic elements:

- The **data**, in this case `mtcars`. This is the first argument of the ggplot function.
- One or more **geometrical elements** or `geom`s, that form the actual visualization. 
- An **aesthetic mapping** or `aes` that maps columns in the data to aspects of the geometry.

In other words, the plot above is a plot of points (i.e. a scatter plot),
whose x and y positions are mapped to the `hp` and `mpg` columns of the data frame `mtcars`

### Make it pretty!

Besides these three basic elements (data, geoms, and aesthetics),
you can tweak almost every aspect of a plot. 
For example, the code below adds an extra mapping, making the colors of points reflect the number of cylinders of the car. 

Besides this, it adds three extra elements:

- A `scale`. By default, ggplot looks at the data and determines a reasonable scale. This default is generally quite good,
  and means that you can quickly make decent graphs. With the `scale_*` commands, you can tweak this scale, for example by 
  changing color values, the range, etc.
- A `ggtitle`. You can add various elements like titles, subtitles, axis labels etc. to the plot to make it more informative
- A `theme`. Themes determine the look and feel of everything around the data, e.g. the plot background, grid lines, fonts, etc. 
  You can create and tweak themes to change the look and feel of plots, for example to match the house style of a company or journal. 

```{r prettyplot, exercise=TRUE, exercise.setup='mtcars'}
ggplot(mtcars, aes(x=hp, y=mpg, color=cyl)) + 
  geom_point() + 
  scale_color_gradient(low="darkgreen", high="darkred") +
  ggtitle("Car horsepower and fuel economy") + 
  theme_minimal()
```

Feel free to play around, e.g. by inverting the x and y in the mapping, changing the colors, adding axis labels (`+ xlab("Horsepower")`), etc. 

## Scatter plots

A very common but quite insightful plot is the scatter plot. It is useful if you want to express the relation between two numeric variables. 

As you saw above, scatter plots use the `geom_point()` geometry. For this plot type, you have to specify and **x** and **y** mapping. 
In addition, you can specify a `color`, `fill`, and `alpha` (transparency) mapping, as well as `size` and `shape`.

For example, the plot below shows the relation between population density and percentage of people aged 65+ in Dutch municipalities.
It also uses point size to indicate total population size, and color to indicate income 
(see e.g. [http://sape.inf.usi.ch/quick-reference/ggplot2/colour](http://sape.inf.usi.ch/quick-reference/ggplot2/colour)) for a reference of color names in ggplot)

```{r point, exercise=TRUE}
library(tidyverse)
demographics <- read_csv('https://raw.githubusercontent.com/vanatteveldt/ccslearnr/master/data/dutch_demographics.csv')
ggplot(demographics,
       aes(x = v57_density, y = c_65plus, size=v01_pop, color=v132_income)) +
  scale_color_gradient(low="coral", high="darkblue") + 
  geom_point(alpha=.6)
```

Note the use of `alpha` to make the points somewhat transparent, making it a bit easier to see when points overlap. 
This is given as a constant value rather than a mapping, since we want all points to have this. 

## Bar plots

(will be added later)

## Line graphs

Line graphs mostly useful to showcase change over time. 
This uses `geom_line`, which is in many ways similar to `geom_point`, but rather than drawing points, it will draw a line between all points.
Besides the `x`, `y`, `color` and `alpha` mapping, it can have a `linetype` (e.g. dashed, dotted, solid) and a `linewidth`.

For example, the following code shows the proportion of the popular vote for each US Presidential candidate since 2000:

```{r line, exercise=TRUE}
library(tidyverse)
elections <- read_csv('https://raw.githubusercontent.com/vanatteveldt/ccslearnr/master/data/US_Elections_years.csv')
ggplot(elections,
       aes(x=year, y=votes, color=party)) + 
  geom_line()
```

To make this graph a bit nicer, let's make some changes:
- The votes are given as a proportion of 0-1, but most people prefer 0-100%. This can be changed in the `scale` by setting a labelling function. 
- This graph uses the default colors for parties, which do not match our normal ideas of which party should have which color.
  We can fix this by using a manual scale for the color: `scale_color_manual`, which allows us to specify exactly which value should get which color. 
- The data for the minor parties was sometimes missing. For example, the Greens were only included in 2000 and 2020. 
  By default, ggplot draws a line between those points, which misleadingly suggests that they had a result in the intervening years as well.
  We fix this using the `complete` function, which inserts missing combinations of `year` and `party`, setting the `vote` variable to zero.
- Finally, let's make the line for the minor parties dotted, using `mutate` to define major, and then using it as the linetype aesthetic, 
  again setting a manual scale to force the FALSE values to be dashed (and removing the guide/legend).


To achieve the last two changes, we need to preprocess the data before the `ggplot` command, using the `complete` and `mutate` functions.
We do this by placing these in a pipeline before `ggplot`, and then of course omitting the data argument from the ggplot call.
(see the handout on summarizing data for an explanation of pipelines, if needed). 

```{r line2-setup}
library(tidyverse)
elections <- read_csv('https://raw.githubusercontent.com/vanatteveldt/ccslearnr/master/data/US_Elections_years.csv')
```
```{r line2, exercise=TRUE}
library(tidyverse)
elections |> 
  complete(year, party, fill=list(votes=0)) |>
  mutate(major = party == "DEMOCRAT" | party == "REPUBLICAN") |>
  ggplot(aes(x=year, y=votes, color=party, linetype=major)) + 
  scale_y_continuous(labels = scales::percent) + 
  scale_linetype_manual(values=c("dashed", "solid"), guide='none') + 
  scale_color_manual(breaks=c("DEMOCRAT", "REPUBLICAN", "GREEN", "LIBERTARIAN", "OTHER"),
                     values=c("blue", "red", "green4", "yellow4", "grey")) + 
  geom_line()
```


## Combining geoms

(will be added later)

## Creating high-quality graphs for reports

(will be added later)
